Comparing R packages for calculation of persistent homology

### 2020-03-19 tdabench-paper/tdabench.Rmd
added line about alpha's limits being computational not mathematical
modified discussion of alpha complex limitation

\section{Methods}
<!--An exclusively R (v3.6.2) code pipeline was built for this project \citep{Rlang,ihaka1996}.-->
We use \CRANpkg{readr} v1.3.1 to read rectangular data \citep{readr}, \CRANpkg{ggplot2} v3.2.1 \citep{ggplot2}, \CRANpkg{scatterplot3d} v0.3-41 \citep{scatterplot3d}, \CRANpkg{recexcavAAR} v0.3.0 \citep{recexcavAAR}, and \CRANpkg{magick} v2.2 \citep{magick} to visualize data, \CRANpkg{bench} v1.0.4 to collect benchmark data \citep{bench}, \CRANpkg{TDA} v1.6.9 \citep{TDA} and \CRANpkg{TDAstats} v0.4.1 \citep{TDAstats} to calculate persistent homology of Vietoris-Rips and alpha simplicial complices, and \CRANpkg{pryr} v0.1.4 for calculations involving R objects \citep{pryr}.
Runtime calculations were averaged over 10 iterations and are visualized as mean $\pm$ SD.
Datasets were generated by sampling functions in base R to generate points uniformly distributed over circles (dimension = 2), spheres (dimension = 3), filled squares (dimension = 2), filled cubes (dimension = 3, 4), and tori (dimension = 3).
The number of points per point cloud varied from 25 to 500 along intervals of 25 points, which were empiric limits chosen after considering available computational resources.
For consistency between software libraries, the minimum and maximum simplicial complex radii were predetermined for each point cloud and provided as parameters to the \CRANpkg{TDA} and \CRANpkg{TDAstats} R packages.
Within the \CRANpkg{TDA} package, benchmark data was collected for the GUDHI \citep{gudhi} and Dionysus \citep{Dionysus} libraries; within the \CRANpkg{TDAstats} package, benchmark data was collected for the Ripser \citep{Ripser} library.
As alpha complex calculation was only implemented in GUDHI, alpha complex benchmark data was naturally only collected for the single library.
Measuring memory usage proved challenging since all the libraries calculating persistent homology were implemented in either C++ or Java and then wrapped in R as part of a CRAN package.
Thus, memory burden was indirectly measured by using boundary matrix size as a proxy.
Given that Ripser optimizes computation of persistent homology by avoiding calculation of a boundary matrix, memory use benchmarks are not provided for Ripser and, consequently, \CRANpkg{TDAstats}.

@@

<!--BELONGS IN DISCUSSION
This is predicted by Vietoris-Rips theoretical exponential complexity and Alpha complex's theoretical polynomial complexity.-->

@@

However, without sufficient computational resources, alpha complexes were not usable for point cloud dimensions greater than 3.
If a point cloud has more than 3 dimensions, then it could undergo pre-processing with dimension reduction before using alpha complexes.
However, if data dimension cannot be reduced without significant information loss, then Vietoris-Rips complexes should be used.
It should be stated that if a point cloud is compatible with both complexes, both analyses should be performed as there may be a variation the persistent homology matrix. 
Among the tested libraries, Ripser (wrapped by \CRANpkg{TDAstats}) has the fastest calculation time. 
GUDHI and Dionysus (wrapped by \CRANpkg{TDA}) significantly fall behind as feature dimension and number of points increase. 
GUDHI and Dionysus, however, do have the option of returning the boundary matrix, which Ripser does not store for the sake of optimization.
While we did not measure Ripser's boundary matrix object size, Ripser's C++ code (outside its implementation in R) demonstrates memory advantages ~\citep{Ripser}

@@

<!--Include somewhere:

differences are b/n implementations (practical) not necessarily mathematical structures (theory), so results can change as implementations change (whether b/n libraries or b/n complices)

Ripser memory not measured here b/c R, but refer to Bauer Ripser memory measurements.

Thoughts on why sphere alpha uses less memory than 3-d annulus alpha

Weigh limitations and benefits of each library/complex type based on Table 1
-->

<!--PUT AS TEXT INSTEAD
\begin{figure}
  \centering
  \includegraphics[height=2.5in]{decision.png}
    \caption{Basic decision chart for choosing TDA library based on dataset.}
\label{decision}
\end{figure}-->

>>>>

\section{Methods}

We use \CRANpkg{readr} v1.3.1 to read rectangular data \citep{readr}, \CRANpkg{ggplot2} v3.2.1 \citep{ggplot2}, \CRANpkg{scatterplot3d} v0.3-41 \citep{scatterplot3d}, \CRANpkg{recexcavAAR} v0.3.0 \citep{recexcavAAR}, and \CRANpkg{magick} v2.2 \citep{magick} to visualize data, \CRANpkg{bench} v1.0.4 to collect benchmark data \citep{bench}, \CRANpkg{TDA} v1.6.9 \citep{TDA} and \CRANpkg{TDAstats} v0.4.1 \citep{TDAstats} to calculate persistent homology of Vietoris-Rips and alpha simplicial complices, and \CRANpkg{pryr} v0.1.4 for calculations involving R objects \citep{pryr}.
Runtime calculations were averaged over 10 iterations and are visualized as mean $\pm$ SD.
Datasets were generated by sampling functions in base R to generate points uniformly distributed over circles (dimension = 2), spheres (dimension = 3), filled squares (dimension = 2), filled cubes (dimension = 3, 4), and tori (dimension = 3).
The number of points per point cloud varied from 25 to 500 along intervals of 25 points, which were empiric limits chosen after considering available computational resources.
For consistency between software libraries, the minimum and maximum simplicial complex radii were predetermined for each point cloud and provided as parameters to the \CRANpkg{TDA} and \CRANpkg{TDAstats} R packages.
Within the \CRANpkg{TDA} package, benchmark data was collected for the GUDHI \citep{gudhi} and Dionysus \citep{Dionysus} libraries; within the \CRANpkg{TDAstats} package, benchmark data was collected for the Ripser \citep{Ripser} library.
As alpha complex calculation was only implemented in GUDHI, alpha complex benchmark data was naturally only collected for the single library.
Measuring memory usage proved challenging since all the libraries calculating persistent homology were implemented in either C++ or Java and then wrapped in R as part of a CRAN package.
Thus, memory burden was indirectly measured by using boundary matrix size as a proxy.
Given that Ripser optimizes computation of persistent homology by avoiding calculation of a boundary matrix, memory use benchmarks are not provided for Ripser and, consequently, \CRANpkg{TDAstats}.

@@

@@

However, without sufficient computational resources, alpha complexes were not usable for point cloud dimensions greater than 3.
If a point cloud has more than 3 dimensions, then it could undergo pre-processing with dimension reduction before using alpha complexes.
Note, it is possible an algorithm will eventually be developed to enable alpha complex computation of higher dimensional data. 
However, if data dimension cannot be reduced without significant information loss, then Vietoris-Rips complexes should be used.
It should be stated that if a point cloud is compatible with both complexes, both analyses should be performed as there may be a variation the persistent homology matrix. 
Among the tested libraries, Ripser (wrapped by \CRANpkg{TDAstats}) has the fastest calculation time. 
GUDHI and Dionysus (wrapped by \CRANpkg{TDA}) significantly fall behind as feature dimension and number of points increase. 
GUDHI and Dionysus, however, do have the option of returning the boundary matrix, which Ripser does not store for the sake of optimization.
While we did not measure Ripser's boundary matrix object size, Ripser's C++ code (outside its implementation in R) demonstrates memory advantages ~\citep{Ripser}

@@
